---
title: "Final_project"
format: html
editor: visual
---

## Loading data

```{r}
data <- read.csv("risk_factors_cervical_cancer.csv")
```

## Loading library

```{r}
library(tidyverse)
library(dplyr)
library(patchwork)
```

## Cleaning

Correcting NAs

```{r}
data <- data %>%
  mutate(across(where(is.character), ~na_if(., "?")))
```

Changing binaries to 'yes' and 'no'

```{r}
data <- data %>%
  mutate(Smokes = case_when(
    Smokes == "0.0" ~ "no",
    Smokes == "1.0" ~ "yes"),
    Smokes = str_to_title(Smokes))

data <- data |> 
  mutate(Hormonal.Contraceptives = case_when(
    Hormonal.Contraceptives == "1.0" ~ "yes",
    Hormonal.Contraceptives == "0.0" ~ "no"),
    Hormonal.Contraceptives = str_to_title(Hormonal.Contraceptives))

data <- data |>  
  mutate(IUD = case_when(IUD == "0.0" ~ "no",
                         IUD == "1.0" ~ "yes"))

data <- data |>
  mutate(Dx.Cancer = case_when(Dx.Cancer == "0" ~ "Negative",
                               Dx.Cancer == "1" ~ "Positive"))

```

Updating names, removing dots

```{r}
data <- data |> 
  rename('Smokes.years'= Smokes..years.,
         'Smokes.packs.years' = Smokes..packs.year.) |> 
  mutate(Smokes.years = as.numeric(Smokes.years),
         Smokes.packs.years = as.numeric(Smokes.packs.years))


data <- data |> 
  rename_with(~ gsub("^STDs\\.\\.", "", .))
```

From character to integer

```{r}
data  <- data|> 
  mutate(Number.of.sexual.partners = as.integer(Number.of.sexual.partners),
         First.sexual.intercourse = as.integer(First.sexual.intercourse),
         Num.of.pregnancies = as.integer(Num.of.pregnancies))
```

Creating a Patient ID

```{r}
data <- data |> 
  mutate(ID = row_number()) |> 
  select(ID, everything())
```

## Creating long data

```{r}
data_long <- data %>%
  pivot_longer(cols = starts_with("STDs."), 
               names_to = "STD_type", 
               values_to = "has_STD") %>%
  mutate(
    STD = ifelse(has_STD == "1.0", STD_type, NA)  # Keep STD name where 1 is present
  ) %>%
  group_by(ID) %>%
  mutate(
    # Concatenate STD names for each ID, if none, set "No"
    STD = ifelse(all(is.na(STD)), "No", paste(na.omit(STD), collapse = ", "))
  ) %>%
  ungroup() %>%
  select(-STD_type, -has_STD) %>%
  distinct() |> 
  separate_rows(STD, sep = ",")
```

## Changing '.' to '\_' in every column. Lastly, we are removing the double underscores if any.

```{r}
data_long <- data_long |> 
  rename_with(~ str_remove(.,"\\.$")) |> #removes the '.' from the last word in columns
  rename_with(~ str_replace_all(., "\\.", "_")) |> 
  rename_with(~ str_replace_all(.x, "__", "_"))
```

## Cleaning up STD

Removed STDs from the beginning of each row in column STD. Capitalize each word.

```{r}
data_long <- data_long |> 
  mutate(STD = str_remove(STD, "^STDs\\."),
         STD = str_remove(STD, "^ STDs\\."),
         STD = str_replace_all(STD, "\\.", " "),
         STD = str_to_title(STD))
```

Remove column STDs (We have the new STD column).

```{r}
data_long <- data_long |> 
  select(-STDs, -number) 
```

Creating new number column that counts number of the same IDs.

```{r}
data_long <- data_long |> 
  group_by(ID) |> 
  mutate(Number_of_STDs = if_else(all(is.na(STD) | STD == "No"), 0, n_distinct(STD, na.rm = TRUE))) |> 
  ungroup() 
```

Moving STD and Number of STD columns further ahead.

```{r}
data_long <- data_long |> 
  select(
    1:12,               
    STD,                
    Number_of_STDs,      
    everything()        
  )
```

<<<<<<< HEAD
## Plotting
=======
>>>>>>> efa4562390202cc5b292e394539a0e2deb5ff205

```{r}
Numb_sex <- ggplot(data = data_long, aes(y = Dx_Cancer, x = Number_of_sexual_partners, fill = Dx_Cancer)) +
  geom_boxplot(show.legend = FALSE) +
   labs(y = "Cancer Diagnosis",
        x = "Number of Sexual Partners") +
  theme_bw()


first_sex_int <- ggplot(data = data_long, aes(y = Dx_Cancer, x = First_sexual_intercourse, fill = Dx_Cancer)) +
  geom_boxplot(show.legend = FALSE) +
   labs(y = "Cancer Diagnosis",
        x = "Age of First Sexual Intercourse") +
  theme_bw()


(Numb_sex / first_sex_int) +
  plot_annotation(
    title = "Comparison of Sexual Health and Cancer Diagnosis",
    caption = "Source: https://doi.org/10.24432/C5Z310")




(Numb_sex / first_sex_int) +
  plot_annotation(
    title = "Comparison of Sexual Health Metrics and Cancer Diagnosis",
    subtitle = "Boxplots showing distribution of sexual health data by cancer diagnosis",
    caption = "Data source: Your dataset"
  )

```

Normalized Counts of STD by Cancer Diagnosis

```{r}
data_normalized_STD <- data_long %>%
  group_by(Dx_Cancer, STD) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(Dx_Cancer) %>%
  mutate(prop = count / sum(count))

ggplot(data = data_normalized_STD, aes(x = STD, y = prop, fill = Dx_Cancer)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_grid(~Dx_Cancer) +
  labs(title = "Normalized Counts of STD by Cancer Diagnosis",
       x = "STD",
       y = "Proportion",
       fill = "Cancer Diagnosis",
       caption = "Source: https://doi.org/10.24432/C5Z310") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Parked for now


data_normalized_HC <- data_long %>%
  group_by(Hormonal_Contraceptives, STD) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(Hormonal_Contraceptives) %>%
  mutate(prop = count / sum(count))

ggplot(data = data_normalized_HC, aes(x = STD, y = prop, fill = Hormonal_Contraceptives)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_grid(~Hormonal_Contraceptives) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Plotting Mathilde

```{r}
data_long %>%
  group_by(STD) |> 
  summarise(STD_count=n()) |> 
  ggplot(aes(x=STD, y=STD_count)) +
  geom_bar(stat = "identity", color = 'black', fill = 'white') +
  geom_text(aes(label=STD_count),vjust = -0.5, size = 3) +
  theme_classic() +
  labs(x = "STD", y = "Count", title = "Counts of Different STDs") +
  theme(axis.text.x = element_text(hjust = 1,angle = 90, vjust = 0.5))
```

\`\`\`
