---
title: "working document"
format: html
editor: visual
---

## Common

```{r}
data <- read.csv("risk_factors_cervical_cancer.csv")
```

```{r}
library(tidyverse)
library(dplyr)
```

## Victor

Changes ? to NAs

```{r}
data <- data %>%
  mutate(across(where(is.character), ~na_if(., "?")))
```

## Nina

Smoking status in column 'Smokes': Change 1.0 and 0.0 to yes and no.

```{r}
data <- data %>%
  mutate(Smokes = case_when(
    Smokes == "0.0" ~ "no",
    Smokes == "1.0" ~ "yes"
  ))
```

## Mariam

Hormonal Contraceptives: Change 1.0 and 0.0 to yes and no.

```{r}
data <- data %>%
  mutate(Hormonal.Contraceptives = case_when(
    Hormonal.Contraceptives == "1.0" ~ "yes",
    Hormonal.Contraceptives == "0.0" ~ "no"))
```

## Mathilde

First, we renamed the two smoke columns to match the other columns syntax. Next, we changed the columns two numeric values.

```{r}
data <- data |> 
  rename('Smokes.years'= Smokes..years.,
         'Smokes.packs.years' = Smokes..packs.year.) |> 
  mutate(Smokes.years = as.numeric(Smokes.years),
         Smokes.packs.years = as.numeric(Smokes.packs.years))
```

Changing the three columns to integers instead of characters.

```{r}
data  <- data|> 
  mutate(Number.of.sexual.partners = as.integer(Number.of.sexual.partners),
         First.sexual.intercourse = as.integer(First.sexual.intercourse),
         Num.of.pregnancies = as.integer(Num.of.pregnancies))
```

### New

Creating consistency in column names by changing '.' to '\_' in every column. Lastly we are removing the double underscores.

```{r}
data_long <- data_long |> 
  rename_with(~ str_remove(.,"\\.$")) |> #removes the '.' from the last word in columns
  rename_with(~ str_replace_all(., "\\.", "_")) |> 
  rename_with(~ str_replace_all(.x, "__", "_"))
```

Removed STDs from the beginning of each row in column STD.

```{r}
data_long <- data_long |> 
  mutate(STD = str_remove(STD, "^STDs\\."),
         STD = str_remove(STD, "^ STDs\\."))
```

Splitted the STDs up to their own row and capitalized each word.

```{r}
data_long <- data_long |> 
  separate_rows(STD, sep = "\\.") |>             
  mutate(
    STD = str_to_title(STD)                      
  )
```

Remove column STDs (We have the new STD column).

```{r}
data_long <- data_long |> 
  select(-STDs, -number) 
```

Creating new number column that counts number of the same IDs.

```{r}
data_long <- data_long |> 
  group_by(ID) |> 
  mutate(Number_of_STDs = n()) 
```

Moving STD and Number of STD columns further ahead.

```{r}
data_long <- data_long |> 
  select(
    1:12,               
    STD,                
    Number_of_STDs,      
    everything()        
  )
```

### Plots

## Fie

```{r}
data <- data |> 
  mutate(ID = row_number()) |> 
  select(ID, everything())

data


data <- data |>  mutate(IUD = case_when(IUD == "0.0" ~ "no",
                                        IUD == "1.0" ~ "yes"))


data <- data |> 
  rename_with(~ gsub("^STDs\\.\\.", "", .))


data_long <- data %>%
  pivot_longer(cols = starts_with("STDs."), 
               names_to = "STD_type", 
               values_to = "has_STD") %>%
  mutate(
    STD = ifelse(has_STD == "1.0", STD_type, NA)  # Keep STD name where 1 is present
  ) %>%
  group_by(ID) %>%
  mutate(
    # Concatenate STD names for each ID, if none, set "No"
    STD = ifelse(all(is.na(STD)), "No", paste(na.omit(STD), collapse = ", "))
  ) %>%
  ungroup() %>%
  select(-STD_type, -has_STD) %>%
  distinct()

data_long <- data_long %>%
  separate_rows(STD, sep = ",")
  
```
