---
title: "working document"
format: html
editor: visual
---

## Common

```{r}
data <- read.csv("risk_factors_cervical_cancer.csv")
```

```{r}
library(tidyverse)
library(dplyr)
```

## Victor

Changes ? to NAs
```{r}
data <- data %>%
  mutate(across(where(is.character), ~na_if(., "?")))
```

## Nina

Smoking status in column 'Smokes': Change 1.0 and 0.0 to yes and no.

```{r}
data <- data %>%
  mutate(Smokes = case_when(
    Smokes == "0.0" ~ "no",
    Smokes == "1.0" ~ "yes"
  ))
```

## Mariam

Hormonal Contraceptives: Change 1.0 and 0.0 to yes and no.

```{r}
data <- data %>%
  mutate(Hormonal.Contraceptives = case_when(
    Hormonal.Contraceptives == "1.0" ~ "yes",
    Hormonal.Contraceptives == "0.0" ~ "no"))
```

## Mathilde

First, we are renaming the two smoke columns to match the others syntax. Next, we change the columns to numeric values.

First, we renamed the two smoke columns to match the other columns syntax. Next, we changed the columns two numeric values.
```{r}
data <- data |> 
  rename('Smokes.years'= Smokes..years.,
         'Smokes.packs.years' = Smokes..packs.year.) |> 
  mutate(Smokes.years = as.numeric(Smokes.years),
         Smokes.packs.years = as.numeric(Smokes.packs.years))
```

Changing the three columns to integers instead of characters.

```{r}
data  <- data |> 
  mutate(Number.of.sexual.partners = as.integer(Number.of.sexual.partners),
         First.sexual.intercourse = as.integer(First.sexual.intercourse),
         Num.of.pregnancies = as.integer(Num.of.pregnancies))
```

## Fie

```{r}
data <- data %>%
  mutate(ID = row_number()) %>%
  select(ID, everything())

data


data <- data %>% mutate(IUD = case_when(IUD == "0.0" ~ "no",
                                IUD == "1.0" ~ "yes"))
```

##End of document
