## Libraries
```{r}
library(tidyverse)
```

## Loading data
```{r}
data_clean <- read_csv("~/projects/project_r4bd/Data/02_data_clean.csv")
```
## Data augmenting

Creating a Patient ID
```{r}
data_clean <- data_clean |> 
  mutate(ID = row_number()) |> 
  select(ID, everything())
```

Creating long data

```{r}
data_long <- data_clean |>
  pivot_longer(cols = starts_with("STDs_"), 
               names_to = "STD_type", 
               values_to = "has_STD") |>
  mutate(
    STD = ifelse(has_STD == 1, STD_type, NA)  # Keep STD name where 1 is present
  ) |>
  group_by(ID) |>
  mutate(
    # Concatenate STD names for each ID, if none, set "No"
    STD = ifelse(all(is.na(STD)), "No", paste(na.omit(STD), collapse = ", "))
  ) |>
  ungroup() |>
  select(-STD_type, -has_STD) |>
  distinct() |> 
  separate_rows(STD, sep = ",")
```

Remove STDs from the beginning of each row in column STD. Capitalize each word and remove the underscores.

```{r}
data_long <- data_long |> 
  mutate(STD = str_remove(STD, "^STDs\\_"),
         STD = str_remove(STD, "^ STDs\\_"),
         STD = str_replace_all(STD, "\\_", " "),
         STD = str_to_title(STD))
```


Remove column STDs (We have the new STD column).

```{r}
data_long <- data_long |> 
  select(-STDs, -number) 
```

Creating new number column that counts number of the same IDs.

```{r}
data_long <- data_long |> 
  group_by(ID) |> 
  mutate(Number_of_STDs = if_else(all(is.na(STD) | STD == "No"), 0, n_distinct(STD, na.rm = TRUE))) |> 
  ungroup() 
```

Moving STD and Number of STD columns further ahead.

```{r}
data_long <- data_long |> 
  select(
    1:12,               
    STD,                
    Number_of_STDs,      
    everything()        
  )
```

Creating new column for PCA analysis
```{r}
data_aug <- data_long |>
  mutate(Cancer = if_else(Dx_Cancer == "yes", 1, 0))
```


## Exporting data
```{r}
write_csv(x = data_aug,
          file = "~/projects/project_r4bd/Data/03_data_aug.csv")
```

